---
title: "Gully Susceptibility using Logistic Regression"
author: "Babatunde Fagbohun"
date: "2024-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(sf)
```

```{r}
library(sp)
```
```{r}
library(tidyverse)
```

```{r}
library(raster)
```

```{r}
library(mapview)
```

```{r}
library(rasterVis)
```


#reading in raster files
```{r}
ndvi <- raster("F:/DA/R/Gully_dataset/reclass_2018_ndv1.tif")
```

```{r}
aspect <- raster("F:/DA/R/Gully_dataset/aspect.tif")
```

```{r}
curvature <- raster("F:/DA/R/Gully_dataset/reclass_curvature.tif")
```

```{r}
elevation <- raster("F:/DA/R/Gully_dataset/reclass_elevationn.tif")
```

```{r}
erosivity <- raster("F:/DA/R/Gully_dataset/reclass_erosivity.tif")
```

```{r}
slope <- raster("F:/DA/R/Gully_dataset/reclass_slope.tif")
```

```{r}
twi <- raster("F:/DA/R/Gully_dataset/Reclass_TWI_filter_2nd1.tif")
```

```{r}
landuse <- raster("F:/DA/R/Gully_dataset/lulc_recl.tif")
```

```{r}
geology <- raster("F:/DA/R/Gully_dataset/reclass_geo.tif")
```

```{r}
soil <- raster("F:/DA/R/Gully_dataset/reclass_soil.tif")
```

#converting raster to a list file type
```{r}
raster_list <-as.list(ndvi, aspect, curvature, elevation, erosivity, slope ,twi, landuse, geology, soil)
```

```{r}
#resampling so that all raster will have same dimension and origin
raster_list_resample <- lapply(raster_list, function(x) resample(x, raster_list[[10]], method='ngb'))
```


#stacking raster files
```{r}
gully_factors_stack <- stack(raster_list_resample, quick=TRUE)
```


#checking number of layers
```{r}
nlayers(gully_factors_stack)
```

```{r}
names(gully_factors_stack)
```

#renaming bands of the raster file
```{r}
names(gully_factors_stack) <- c("ndvi", "aspect", "curvature","elevation", "erosivity", "slope", "twi", "landuse", "geology", "soil")
```

```{r}
library(RColorBrewer)
```

```{r}
MyColour <- brewer.pal(11, "RdYlGn")
```

```{r}
MyColour
```

```{r}
MyPalette<-colorRampPalette(MyColour)
```

```{r}
plot(gully_factors_stack, col=rev(MyColour))
```

#reading in shapefile of gully points and non-gully points
```{r}
gully_pts <- st_read("F:/DA/R/Gully_dataset/gully_tr_pts.shp", quiet = T)
```

```{r}
non_gully_pts <- st_read("F:/DA/R/Gully_dataset/non_gully_tr_pts.shp", quiet = T)
```


#reading in shapefile of gully points and non-gully points
```{r}
summary(gully_pts)
```

```{r}
summary(non_gully_pts)
```


```{r}
library(terra)
```

```{r}
library(ggplot2)
```

```{r}
library(terra)
```

```{r}
library(tidyterra)
```

#plotting the distribution of training points
```{r}
ggplot()+
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  geom_sf (data=gully_pts, color="black")+
  geom_sf(data=non_gully_pts, color="grey")+
  ggtitle ("Distribution of training points")+
  coord_sf()
```

#converting ndvi to terra raster 
```{r}
ndvi_t <-rast(ndvi)
```



#plotting the distribution of training points over one raster
```{r}
ggplot()+
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  geom_spatraster(data=ndvi_t)+
  geom_sf (data=gully_pts, aes(col="red"))+
  geom_sf(data=non_gully_pts, aes(col="lightblue"))+
  scale_fill_gradientn(colours=MyPalette(20), na.value=NA)+
  scale_color_identity(labels=c(red="Gully", lightblue="Non-gully"), guide="legend")+
  ggtitle ("distribution of training points")+
  coord_sf()
```

#extracting the attribute presence/absence from the imported shapefiles
```{r}
gully <-gully_pts$gng
```

```{r}
non_gully <- non_gully_pts$gng
```


#converting imported shapefile to a spatial feature
```{r}
g_pts_sp <-as(gully_pts, 'Spatial')
```

```{r}
ng_pts_sp <-as(non_gully_pts, 'Spatial')
```

#extracting values from the stacked raster file to the shape file of gully occurrences
```{r}
gully_ext <- raster::extract(gully_factors_stack, g_pts_sp, method="simple")
```

```{r}
head(gully_ext)
```

```{r}
tail(gully_ext)
```

```{r}
summary(gully_ext)
```

#extracting values from the stacked raster file to the shape file of non-gully points
```{r}
ng_ext <- raster::extract(gully_factors_stack, ng_pts_sp, method="simple")
```

```{r}
head(ng_ext)
```


#appending presence/absence (1/0) information to the data extracted from the raster
```{r}
g_ext_df <- data.frame(gully_ext, gully)
```

```{r}
ng_ext_df <- data.frame(ng_ext, non_gully)
```

#naming the column
```{r}
colnames(g_ext_df)[11] <- "presence_absence"
```

```{r}
colnames(ng_ext_df)[11] <- "presence_absence"
```

#combining extracted data for gully points and non-gully points
```{r}
training_pts <- rbind(g_ext_df, ng_ext_df)
```

```{r}
head(training_pts)
```

```{r}
summary(training_pts)
```

```{r}
names(training_pts)
```



```{r}
par(mfrow=c(2,5))
```


#checking the distribution of continuous factors in both gully and non-gully points
```{r}
boxplot(aspect ~ presence_absence, data=training_pts, main= "Aspect", xlab=" ", ylab="Aspect", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```

```{r}
boxplot(ndvi ~ presence_absence, data=training_pts,main= "NDVI", xlab=" ", ylab=" ", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```

```{r}
boxplot(curvature ~ presence_absence, data=training_pts, main="Curvature", xlab=" ", ylab="Curvature", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```

```{r}
boxplot(elevation ~ presence_absence, data=training_pts, main="Elevation", xlab=" ", ylab="Elevation", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```

```{r}
boxplot(erosivity ~ presence_absence, data=training_pts, main="Erosivity", xlab="Erosivity", ylab="", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```

```{r}
boxplot(slope ~ presence_absence, data=training_pts, main="Slope", xlab=" ", ylab="Slope", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```
```{r}
boxplot(twi ~ presence_absence, data=training_pts, main="TWI", xlab=" ", ylab=" ", boxwex=0.3, col="lightblue", border="darkblue", names=c("Non-gully", "Gully"))
```


```{r}
par(mfrow=c(1,1))
```

#checking the number of samples in class i.e gully and non-gully points
```{r}
aggregate(training_pts$soil ~training_pts$presence_absence, FUN=length)
```


#remove few NAs introduced due to points at the edges study area and duplicated training data
```{r}
training_pts1 <- na.omit(training_pts)
```


#converting categorical data to factor data type
```{r}
training_pts1$landuse <-as.factor(training_pts1$landuse)
```

```{r}
training_pts1$geology <-as.factor(training_pts1$geology)
```

```{r}
training_pts1$soil <-as.factor(training_pts1$soil)
```


```{r}
levels(training_pts1$geology)
```


```{r}
library(caret)
```


#implement one hot coding for categorical data
```{r}
dummy <- dummyVars(" ~ .", data=training_pts1)
```

```{r}
training_df <- data.frame(predict(dummy, newdata=training_pts1))
```

```{r}
training_df %>% 
  head(n=3)
```
```{r}
names(training_df)
```



#Split data into training and test
```{r}
training_df$random<-runif(length(training_df$presence_absence))
```

#runif function provides information about the uniform distribution on the interval
#from min to max, to generate random deviates.
```{r}
train<-training_df[training_df$random<0.70, 1:23] ## the training data
```

```{r}
evalu<-training_df[training_df$random>=0.70, 1:23] ## the evaluation data
```

```{r}
nrow(train)
```

```{r}
nrow(evalu)
```

```{r}
names(train)
```

```{r}
names(evalu)
```


```{r}
library(MASS)
```

```{r}
log_model <- suppressWarnings(glm(presence_absence~., data=train, family="binomial")%>% stepAIC(trace=FALSE))
```

```{r}
summary(log_model)
```


#Predicting the probabilities for test data
```{r}
evalu$lr_pred <-log_model%>% predict(evalu, type="response")
```

```{r}
summary(evalu)
```


#Accuracy assessment through AUC/ROC Plot
```{r}
library(PresenceAbsence)
```

```{r}
pa_validate<-data.frame(ID=1:length(evalu$presence_absence),
                        PA=evalu$presence_absence,
                        logistic=evalu$lr_pred)
```

```{r}
presence.absence.accuracy(pa_validate, threshold=0.5)
```

```{r}
error.threshold.plot(pa_validate,which.model=1)
```

```{r}
auc.roc.plot(pa_validate)
```

#Accuracy assessment through comparison of mean and confusion matrix
#Predictions values <0.5 are converted to 0, >0.5 are converted 1
```{r}
evalu$lr_pred1<-ifelse(evalu$lr_pred>0.5, "1", "0")
```

```{r}
mean(evalu$presence_absence==evalu$lr_pred1)
```

```{r}
confusionMatrix((as.factor(evalu$presence_absence)), (as.factor(evalu$lr_pred1)))
```


#converting the raster stack to a dataframe
```{r}
raster_df <-terra::as.data.frame(gully_factors_stack, xy=TRUE)
```

```{r}
raster_df$landuse <-as.factor(raster_df$landuse)
```

```{r}
raster_df$geology <-as.factor(raster_df$geology)
```

```{r}
raster_df$soil <-as.factor(raster_df$soil)
```


#applying one hot coding to categorical data
```{r}
dummy <- dummyVars(" ~ .", data=raster_df)
```

```{r}
raster_df <- data.frame(predict(dummy, newdata=raster_df))
```


#making prediction for the converted dataframe created from raster files
```{r}
raster_df$logreg <- log_model %>% predict(raster_df[,3:23], type="response")
```

```{r}
head(raster_df)
```


#converting the data frame back to raster
```{r}
gully_ras_logreg <- rasterFromXYZ(raster_df)
```

```{r}
projection(gully_ras_logreg) <-projection(gully_factors_stack)
```



#Minimum and maximum prediction values
```{r}
min(gully_ras_logreg$logreg@data@values, na.rm = T)
```

```{r}
cellStats(gully_ras_logreg$logreg, max)
```


#Plot the gully susceptibility map
```{r}
plot(gully_ras_logreg$logreg, main="Gully Susceptibility Map", col=rev(MyPalette(20)))
```




